Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-01-27 draft.tex
.

\section{Introduction}
\subsection{Homeostasis in the Nervous System}
Among other functional properties of the mammalian central nervous system, it is involved in maintaining crucial properties of the body in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. In more recent years, experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}. Many processes underlying this self-regulatory behavior have been found up till now, and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though, experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

@@

Theoretical models of homeostasis are generally based on a form of feedback control. As such, it necessarily contains an error estimation. This could be based e.g. on the neuron's firing rate of the recent past measured in a given window, or, on a physiological level, e.g. a concentration of some molecule within some part of the cell structure. In any case, the error estimation requires setting some predefined target that the feedback control will attempt to achieve. This raises two questions: First, what would be an appropriate choice of homeostatic targets? Second, given this set of targets, how could a neural system with fixed homeostatic targets encode changes of external stimuli if homeostasis overrides any information about changes in synaptic input?

A possible way of choosing homeostatic targets could be to randomly draw from a distribution that matches statistics known to be found in the neural tissue that one wishes to model. This, however is a highly artificial ad-hoc solution. Furthermore, it does not resolve the second issue of encoding of external input.

@@

\subsection{Current State of the Network Model}
This thesis is based on earlier versions of a self-organizing recurrent neural network (SORN), in particular an implementation using leaky integrate-and-fire neurons (LIF-SORN) \cite{SORN_Paper}. It was developed to model features of activity and topology of cortical networks, in particular L5 of the neocortex. In these previous implementations, a slow homeostatic mechanism was implemented among excitatory neurons by an intrinsic adaptation of excitability. This led to the aforementioned problems of homeostatic control: Neural activity was tuned to a globally set target, allowing for no variability. In contrast, cortical networks are known to exhibit a broad distribution of firing rates, which is well described by a right-skewed, log-normal like shape \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Thus, if the LIF-SORN was meant to provide a good model for cortical structure \emph{and} activity, it would require a different type of homeostatic model.

@@

Apart from reproducing the desired connection fraction, previous versions of the LIF-SORN showed an over-representation of bidirectional connections compared to a random graph with equal connectivity \cite{SORN_Paper}. This experimentally observed feature \cite{Markram_Connections_1997,Song_Connectivity_2005} required the presence of a distance dependent connection probability, as described in section \ref{network simulation}. The absence of this breaking of spatial homogeneity even led to an under-representation of bidirectional connections, which is known to be an effect of STDP in recurrent networks \cite{Syn_Plast_Abbott}. As shown in Figure \ref{RC_CFB_plot}, the separation between simulations with and without a spatial connection profile is retained. 

>>>>

\section{Introduction}
\subsection{Homeostasis in the Nervous System}
Among other functional roles of the mammalian central nervous system, it is involved in maintaining crucial properties of the body in a functional regime. Some examples regarding the organism as a whole are e.g. body temperature and blood pressure, both being controlled by hypothalamus \cite[p. 484]{Bear_Exploring_the_Brain}. The term \textit{homeostasis} entailing these regulatory mechanisms was first introduced by Walter C. Cannon in 1932 \cite{Wisdom_of_the_Body_Cannon, Homeostasis_Queenan_2012}. In more recent years, experimental studies have found evidence for processes within the brain that do not function as a regulatory control for certain aspects of the body, but as a homeostatic control of neural activity itself \cite{Turrigiano_1994,Burrone_2003}. Many processes underlying this self-regulatory behavior have been found up till now, and many different biophysical interdependencies have shown to play a role. This diversity makes comparability of these mechanisms a rather difficult task. On a relatively general level though, experiments have revealed two main types of neuronal regulation: synaptic homeostatic plasticity and intrinsic homeostasis \cite{Desai_2003}.

@@

Theoretical models of homeostasis are generally based on a form of feedback control. As such, they necessarily contain an error estimation. This could be based e.g. on the neuron's firing rate of the recent past measured in a given window, or, on a physiological level, e.g. a concentration of some molecule within some part of the cell structure. In any case, the error estimation requires setting some predefined target that the feedback control will attempt to achieve. This, of course, raises the question of what would be an appropriate choice of homeostatic targets.

Experimental studies have found that neuronal populations exhibit a strong heterogeneity within their individual activity. In particular, many statistical analyses found evidence for a log-normal like distribution of firing rates \cite{Buzsaki_Fir_Rates_2014,Wohrer_Fir_Rates_2012}. Following these results, one approach of modeling this feature would be to randomly distribute individual targets according to the aforementioned statistics. A problem of this procedure is the absence of any particular deeper physiological motivation or causation. In the light of this conceptual problem, one could, due to a lack of better knowledge, simply define a single global target of activity. However, it has been shown that heterogeneity of neural response onto external input within a population is crucial for efficient stimulus encoding \cite{Marsat_2010}. Allowing for a broad distribution of activity thus has to be regarded an important compartment of models of cortical activity and structure.

\subsection{Earlier Research on the Network Model}
This thesis is based on earlier versions of a self-organizing recurrent neural network (SORN), in particular an implementation using leaky integrate-and-fire neurons (LIF-SORN) \cite{SORN_Paper}. It was developed to model features of activity and topology of cortical networks, in particular L5 of the neocortex. In these previous implementations, a slow homeostatic mechanism was implemented among excitatory neurons by an intrinsic adaptation of excitability. This led to the aforementioned problems of homeostatic control: Neural activity was tuned to a globally set target, allowing for no variability. Despite this limitation, the model has proven to be capable of showing a number of experimentally confirmed features. Zheng et al. have shown that distribution and dynamics of synaptic efficacies measured in rat hippocampus can be reproduced by a binary SORN using a discretized version of spike-timing-dependent plasticity and presynaptic normalization \cite{Pengsheng_2013}. Based on this model, Miner et al. designed the LIF-SORN. This network reproduced the results of the binary SORN regarding self-organizing synaptic dynamics. Furthermore, it was able to explain the experimentally observed overrepresentation of bidirectional connections (see \cite{Markram_Connections_1997,Song_Connectivity_2005}) by means of a distance-dependent implementation of synaptic growth.

@@

@@

Apart from reproducing the desired connection fraction, previous versions of the LIF-SORN showed an over-representation of bidirectional connections compared to a random graph with equal connectivity \cite{SORN_Paper}. This experimentally observed feature required the presence of a distance dependent connection probability, as described in section \ref{network simulation}. The absence of this breaking of spatial homogeneity even led to an under-representation of bidirectional connections, which is known to be an effect of STDP in recurrent networks \cite{Syn_Plast_Abbott}. As shown in Figure \ref{RC_CFB_plot}, the separation between simulations with and without a spatial connection profile is retained. 
